# Impact of discrepant reference annotations on scRNA-seq analyses

In this section, we will have a look at the differences between the three references genome annotation: Ensembl, NCBI (RefSeq) and UCSC.

## Discrepancies in reference annotations

### Define paths and get reference annotations

These are just sanity steps to define the inputs and outputs of the workflow.

```{bash, eval=TRUE}
echo "export pathRef='data/raw/references/annotations/ucsc/'" >> .rvars
echo "export pathOutput='output/01-Impact-ref-annotation-scRNA/'" >> .rvars
```

```{bash, eval=FALSE, message=FALSE, warning=FALSE}
source .rvars
mkdir -p $pathRef
mkdir -p $pathOutput

wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.ensGene.gtf.gz \
-P $pathRef
wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.ncbiRefSeq.gtf.gz \
-P $pathRef
wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.refGene.gtf.gz \
-P $pathRef
gunzip ${pathRef}*.gtf.gz
```

```{bash, eval=TRUE, echo=FALSE}
source .rvars
echo "export ref1='${pathRef}galGal6.ensGene.gtf'" >> .rvars
echo "export ref2='${pathRef}galGal6.ncbiRefSeq.gtf'" >> .rvars
echo "export ref3='${pathRef}galGal6.refGene.gtf'" >> .rvars
```

### Basic stats on GTF files

We here show some basic statistics out of the genome annotation file.
For this purpose, we use [mikado util stats](https://mikado.readthedocs.io/en/stable/Usage/Utilities/#stats).
However, while running it on UCSC reference, we encounter some errors in the original file.
We will start by fixing the file.

#### Fix GTF RefGene file

When running `mikado`, we found out that some exons overlap.
This causes `mikado` to throw an error such as "start must be less than end", and it stops running.
We use `gffread -T` to fix this issue.

```{bash, eval=FALSE}
source .rvars
gffread -T $ref3 > ${ref3%.*}_fixed.gtf
```

```{bash, eval=FALSE, echo=FALSE}
sed -i '/ref3/c\' .rvars
echo "export ref3='${ref3%.*}_fixed.gtf'" >> .rvars
source .rvars
echo $ref3
```

### Run nextflow pipeline

We designed a scAnnotatiONT pipeline in order to process single-cell RNA-seq from poorly-annotated genomes.
This pipeline contains also an option to process reference annotations.
It outputs i) basic statistics on any GTF file, ii) single-cell RNA-seq analyses of our test dataset.

```{bash, eval=FALSE}
# Pipeline available here (see "paper" branch) : https://github.com/LehmannN/scAnnotatiONT
bash _run_nextflow_references.sh
```

### Mikado

```{r mikado-stats}
file_names <- c('ensGene', 'ncbiRefSeq', 'refGene')
ref_names <- c('Ensembl', 'NCBI', 'UCSC')
mikado_files <- c(paste0('output/01-Impact-ref-annotation-scRNA/mikado_stats/galGal6.', file_names,'_stats.tsv'))
gtf_stats <- list()
gtf_stats <- lapply(mikado_files, read.csv,
					header = TRUE,
					sep = '\t',
					row.names = 1,
					quote = "")

show_df <- function(x) {
	DT::datatable(x,
				  extensions = c('Buttons', 'Scroller'),
				  options = list(
								 dom = 'Bfrtip',
								 buttons = c('csv', 'pdf'),
								 deferRender = TRUE,
								 scrollY = 200,
								 scrollX = TRUE,
								 scroller = TRUE))}

show_df(gtf_stats[[1]])
show_df(gtf_stats[[2]])
show_df(gtf_stats[[3]])
```


We also use `asm_collect.py`, which is a script to collect info from multiple mikado util stats files.
See https://mikado.readthedocs.io/en/stable/Usage/Utilities/#stats for more details.

```{bash, eval=FALSE}
asm_collect.py ${pathOutput}*_stats.tsv > ${pathOutput}main_stats.tsv
```

```{r mikado-general-stats}
main_stats <- read.csv("output/01-Impact-ref-annotation-scRNA/mikado_stats/main_stats.tsv",
					header = TRUE,
					sep = '\t',
					row.names = 1,
					quote = "")

show_df(main_stats)
```

### Gene lengths

We extract *genes* features from the GFF3 file (see workflow for more details: GTF files do not include the *genes* features).
This allow us to get gene lengths distribution for each annotation.

```{r gene-length}
gff3_files <- c(paste0('data/raw/references/annotations/ucsc/galGal6.', file_names,'.gff3'))
gff3_list <- list()
gff3_list <- lapply(gff3_files, rtracklayer::import)
gff3_list <- lapply(gff3_list, as.data.frame)
genes <- lapply(gff3_list, function(x) filter(x, type == "gene"))
genes_width <- list(Ensembl = genes[[1]]['width'],
					NCBI = genes[[2]]['width'],
					UCSC = genes[[3]]['width'])
summ <- lapply(genes_width, summary)
knitr::kable(summ[[1]])
knitr::kable(summ[[2]])
knitr::kable(summ[[3]])
```

```{r}
# Define colors
plots_colors <- brewer.pal(8, "Set2")
# Prepare data for plotting
width1 <- data.frame(annotation = "Ensembl", value = genes[[1]]['width'])
width2 <- data.frame(annotation = "NCBI", value = genes[[2]]['width'])
width3 <- data.frame(annotation = "UCSC", value = genes[[3]]['width'])
genes_width_df <- rbind(width1, width2, width3)
```

```{r gene-length-boxplot}
# Boxplot
plot1 <- ggplot(genes_width_df, aes(x = annotation, y = log10(width))) +
	geom_boxplot(width = 0.5, fill = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r gene-length-density}
# Density plot
plot1 <- ggplot(genes_width_df, aes(x = log10(width), color = annotation)) +
	geom_density() +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r gene-length-histogram}
# Histogram
plot1 <- ggplot(genes_width_df, aes(x = log10(width), fill = annotation)) +
	geom_histogram(bins = 30, color = "#e9ecef", alpha = 0.5, position = 'identity') +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```


### 3'UTR lengths

```{r utr-length}
gtf_files <- c(paste0('data/raw/references/annotations/ucsc/galGal6.', file_names,'.gtf'))
gtf_list <- list()
gtf_list <- lapply(gtf_files, rtracklayer::import)
gtf_list <- lapply(gtf_list, as.data.frame)
utr <- lapply(gtf_list, function(x) filter(x, type == "3UTR"))
utr_width <- list(Ensembl = utr[[1]]['width'],
				  NCBI = utr[[2]]['width'],
				  UCSC = utr[[3]]['width'])
summ <- lapply(utr_width, summary)
knitr::kable(summ[[1]])
knitr::kable(summ[[2]])
knitr::kable(summ[[3]])
```

```{r utr-length-boxplot}
# Prepare data for plotting
width1 <- data.frame(annotation = "Ensembl", value = utr[[1]]['width'])
width2 <- data.frame(annotation = "NCBI", value = utr[[2]]['width'])
width3 <- data.frame(annotation = "UCSC", value = utr[[3]]['width'])
utr_width_df <- rbind(width1, width2, width3)

plot1 <- ggplot(utr_width_df, aes(x = annotation, y = log10(width))) +
	geom_boxplot(width = 0.5, fill = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r utr-length-density}
# Density plot
plot1 <- ggplot(utr_width_df, aes(x = log10(width), color = annotation)) +
	geom_density() +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r utr-length-histogram}
# Histogram
plot1 <- ggplot(utr_width_df, aes(x = log10(width), fill = annotation)) +
	geom_histogram(bins = 30, color = "#e9ecef", alpha = 0.5, position = 'identity') +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

### Venn Diagrams
Intersections counts are produced via [Intervene](https://intervene.readthedocs.io/en/latest/index.html).
This tool is based on `bedtools interset`.
We defined intersections if at least 50% of the gene from reference A overlaps a gene in reference B, and vice versa.
We also consider separately the forward and reverse strand.
The resulting euler diagram (or proportional Venn diagram) is obtained with the package `eulerr`.

```{bash, eval=FALSE, echo=FALSE}
source .rvars
sed -i '/pathOutput/c\' .rvars
echo "export pathOutput='output/01-Impact-ref-annotation-scRNA/venn_diagrams/'" >> .vars
source .rvars
mkdir -p $pathOutput
```

```{bash, eval=FALSE, echo=FALSE}
source .rvars
echo "export genesGff1='${pathRef}galGal6.ensGene_genes.gff3'" >> .rvars
echo "export genesGff2='${pathRef}galGal6.ncbiRefSeq_genes.gff3'" >> .rvars
echo "export genesGff3='${pathRef}galGal6.refGene_fixed_genes.gff3'" >> .rvars
```

```{bash, eval=FALSE}
source .rvars

intervene venn -i $genesGff1 $genesGff2 $genesGff3 \
--figtype png \
--save-overlaps \
--bedtools-options f=0.5,r,s \
--output ${pathOutput}
```

```{r venn-refs, eval=TRUE}
venn_intersect <- euler(c("UCSC" = 248,
						  "NCBI" = 8555,
						  "Ensembl" = 9004,
						  "NCBI&Ensembl" = 8927,
						  "NCBI&UCSC" = 253,
						  "Ensembl&UCSC" = 376,
						  "NCBI&Ensembl&UCSC" = 6049))

plot(venn_intersect,
	 quantities = list(type = c("counts"),
					   col =  c("black", "black", "black"),
					   fontsize = 20),
	 labels = list(col = c("black", "black", "black"),
				   fontsize = 22),
	 edges = list(col = "white", lex = 2),
	 fills = plots_colors[3:1],
	 legend = list(side = "right"))
```

### GffCompare

## Impact of reference choice on scRNA-seq analyses

```{r, eval=FALSE}

```
