---
title: "scAnnotatiONT paper companion"
author: "Nathalie Lehmann"
date: "`r Sys.Date()`"
output: bookdown::gitbook
documentclass: book
bibliography: []
biblio-style: apalike
link-citations: yes
github-repo: "https://github.com/LehmannN/scAnnotatiONT"
---


```{r setup-paths, echo=FALSE}
path_main <- './'
path_data <- paste0(path_main, 'data/')
path_output <- paste0(path_main, 'output/')
path_scripts <- paste0(path_main,'scripts/')
```

```{r setup-chunks, echo=FALSE, message=FALSE}
library(knitr)

# Change cache to TRUE if you need to re-run the analysis with small changes
# Careful: Bookdown doesn’t support proper cache invalidation based on external file changes
opts_chunk$set(fig.align = "center",
			   fig.retina = 2,
			   fig.width = 10,
			   dev = c('png', 'pdf'),
			   cache = FALSE,
			   cache.lazy = FALSE,
			   warning = FALSE,
			   message = FALSE)
```

```{r def-libraries, echo=FALSE, cache=FALSE}
packagesList <- c("bookdown",
				  "dplyr",
				  "eulerr",
				  "RColorBrewer",
				  "wesanderson",
				  "ggplot2",
				  "gridExtra",
				  "ggpubr",
				  "Seurat",
				  "gridExtra",
				  "cowplot",
				  "scales",
				  "ggthemes",
				  "ggrepel")
lapply(packagesList, require, character.only = TRUE)
options(stringsAsFactors = FALSE)
```

```{r setup-knitr, include=FALSE}
# Automatically creates a bib database for R packages
knitr::write_bib(c(
				 .packages(), packagesList),
				 'env/packages.bib')
# Adds an option to show / hide the code dynamically in the report
#knitr::read_chunk('src/R/rmd_code_folding.R')
```

```{r ggplot-theme, echo=FALSE}
#Set ggplot theme
#theme_set(theme_classic())
```


<!--chapter:end:00-Index.Rmd-->

# Impact of discrepant reference annotations on scRNA-seq analyses

In this section, we will have a look at the differences between the three references genome annotation: Ensembl, NCBI (RefSeq) and UCSC.

## Discrepancies in reference annotations

### Define paths and get reference annotations

These are just sanity steps to define the inputs and outputs of the workflow.

```{bash, eval=TRUE}
echo "export pathRef='data/raw/references/annotations/ucsc/'" >> .rvars
echo "export pathOutput='output/01-Impact-ref-annotation-scRNA/'" >> .rvars
```

```{bash, eval=FALSE, message=FALSE, warning=FALSE}
source .rvars
mkdir -p $pathRef
mkdir -p $pathOutput

wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.ensGene.gtf.gz \
-P $pathRef
wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.ncbiRefSeq.gtf.gz \
-P $pathRef
wget https://hgdownload.soe.ucsc.edu/goldenPath/galGal6/bigZips/genes/galGal6.refGene.gtf.gz \
-P $pathRef
gunzip ${pathRef}*.gtf.gz
```

```{bash, eval=TRUE, echo=FALSE}
source .rvars
echo "export ref1='${pathRef}galGal6.ensGene.gtf'" >> .rvars
echo "export ref2='${pathRef}galGal6.ncbiRefSeq.gtf'" >> .rvars
echo "export ref3='${pathRef}galGal6.refGene.gtf'" >> .rvars
```

### Basic stats on GTF files

We here show some basic statistics out of the genome annotation file.
For this purpose, we use [mikado util stats](https://mikado.readthedocs.io/en/stable/Usage/Utilities/#stats).
However, while running it on UCSC reference, we encounter some errors in the original file.
We will start by fixing the file.

#### Fix GTF RefGene file

When running `mikado`, we found out that some exons overlap.
This causes `mikado` to throw an error such as "start must be less than end", and it stops running.
We use `gffread -T` to fix this issue.

```{bash, eval=FALSE}
source .rvars
gffread -T $ref3 > ${ref3%.*}_fixed.gtf
```

```{bash, eval=FALSE, echo=FALSE}
sed -i '/ref3/c\' .rvars
echo "export ref3='${ref3%.*}_fixed.gtf'" >> .rvars
source .rvars
echo $ref3
```

### Run nextflow pipeline

We designed a scAnnotatiONT pipeline in order to process single-cell RNA-seq from poorly-annotated genomes.
This pipeline contains also an option to process reference annotations.
It outputs i) basic statistics on any GTF file, ii) single-cell RNA-seq analyses of our test dataset.

```{bash, eval=FALSE}
# Pipeline available here (see "paper" branch) : https://github.com/LehmannN/scAnnotatiONT
bash _run_nextflow_references.sh
```

### Mikado

```{r mikado-stats}
file_names <- c('ensGene', 'ncbiRefSeq', 'refGene')
ref_names <- c('Ensembl', 'NCBI', 'UCSC')
mikado_files <- c(paste0('output/01-Impact-ref-annotation-scRNA/mikado_stats/galGal6.', file_names,'_stats.tsv'))
gtf_stats <- list()
gtf_stats <- lapply(mikado_files, read.csv,
					header = TRUE,
					sep = '\t',
					row.names = 1,
					quote = "")

show_df <- function(x) {
	DT::datatable(x,
				  extensions = c('Buttons', 'Scroller'),
				  options = list(
								 dom = 'Bfrtip',
								 buttons = c('csv', 'pdf'),
								 deferRender = TRUE,
								 scrollY = 200,
								 scrollX = TRUE,
								 scroller = TRUE))}

# Ensembl
show_df(gtf_stats[[1]])
# NCBI
show_df(gtf_stats[[2]])
# UCSC
show_df(gtf_stats[[3]])
```


We also use `asm_collect.py`, which is a script to collect info from multiple mikado util stats files.
See https://mikado.readthedocs.io/en/stable/Usage/Utilities/#stats for more details.

```{bash, eval=FALSE}
asm_collect.py ${pathOutput}*_stats.tsv > ${pathOutput}main_stats.tsv
```

```{r mikado-general-stats}
main_stats <- read.csv("output/01-Impact-ref-annotation-scRNA/mikado_stats/main_stats.tsv",
					   header = TRUE,
					   sep = '\t',
					   row.names = 1,
					   quote = "")

show_df(main_stats)
```

### Gene lengths

```{r extract-gff3}
# show how we created gff3 and why
```


We extract *genes* features from the GFF3 file (see workflow for more details: GTF files do not include the *genes* features).
This allow us to get gene lengths distribution for each annotation.


```{r gene-length}
gff3_files <- c(paste0('data/raw/references/annotations/ucsc/galGal6.', file_names,'.gff3'))
gff3_list <- list()
gff3_list <- lapply(gff3_files, rtracklayer::import)
gff3_list <- lapply(gff3_list, as.data.frame)
genes <- lapply(gff3_list, function(x) dplyr::filter(x, type == "gene"))
genes_width <- list(Ensembl = genes[[1]]['width'],
					NCBI = genes[[2]]['width'],
					UCSC = genes[[3]]['width'])
summ <- lapply(genes_width, summary)

# Ensembl
knitr::kable(summ[[1]])
# NCBI
knitr::kable(summ[[2]])
# UCSC
knitr::kable(summ[[3]])
```

```{r}
# Define colors
plots_colors <- brewer.pal(8, "Set2")
# Prepare data for plotting
width1 <- data.frame(annotation = "Ensembl", value = genes[[1]]['width'])
width2 <- data.frame(annotation = "NCBI", value = genes[[2]]['width'])
width3 <- data.frame(annotation = "UCSC", value = genes[[3]]['width'])
genes_width_df <- rbind(width1, width2, width3)
```

```{r gene-length-boxplot}
# Boxplot
plot1 <- ggplot(genes_width_df, aes(x = annotation, y = log10(width))) +
	geom_boxplot(width = 0.5, fill = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r gene-length-significance}
comp_genes_len <- compare_means(data = genes_width_df,
								formula = width ~ annotation,
								method = "wilcox.test", paired = FALSE)
show_df(as.data.frame(comp_genes_len))
```

Comparison of mean gene lengths.
```{r gene-length-boxplot-advanced}
to_compare <- list(c('Ensembl', 'NCBI'), c('NCBI', 'UCSC'), c('Ensembl', 'UCSC'))
plot1 <- ggplot(genes_width_df, aes(x = annotation, y = log10(width))) +
	geom_violin(trim = TRUE, fill = 'lightgray', color = "lightgray") +
	geom_boxplot(width = 0.4, fill = plots_colors[1:3]) +
	stat_compare_means(comparisons = to_compare) +
	stat_compare_means(label.y = 11) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```


```{r gene-length-density}
# Density plot
plot1 <- ggplot(genes_width_df, aes(x = log10(width), color = annotation)) +
	geom_density() +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r gene-length-histogram}
# Histogram
plot1 <- ggplot(genes_width_df, aes(x = log10(width), fill = annotation)) +
	geom_histogram(bins = 30, color = "#e9ecef", alpha = 0.5, position = 'identity') +
	scale_fill_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("Genes lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```


### 3'UTR lengths

```{r utr-length}
gtf_files <- c(paste0('data/raw/references/annotations/ucsc/galGal6.', file_names,'.gtf'))
gtf_list <- list()
gtf_list <- lapply(gtf_files, rtracklayer::import)
gtf_list <- lapply(gtf_list, as.data.frame)
utr <- lapply(gtf_list, function(x) filter(x, type == "3UTR"))
utr_width <- list(Ensembl = utr[[1]]['width'],
				  NCBI = utr[[2]]['width'],
				  UCSC = utr[[3]]['width'])
summ <- lapply(utr_width, summary)
# Ensembl
knitr::kable(summ[[1]])
# NCBI
knitr::kable(summ[[2]])
# UCSC
knitr::kable(summ[[3]])
```

```{r utr-length-boxplot}
# Prepare data for plotting
width1 <- data.frame(annotation = "Ensembl", value = utr[[1]]['width'])
width2 <- data.frame(annotation = "NCBI", value = utr[[2]]['width'])
width3 <- data.frame(annotation = "UCSC", value = utr[[3]]['width'])
utr_width_df <- rbind(width1, width2, width3)

plot1 <- ggplot(utr_width_df, aes(x = annotation, y = log10(width))) +
	geom_boxplot(width = 0.5, fill = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r utr-length-significance}
comp_utr_len <- compare_means(data = utr_width_df,
							  formula = width ~ annotation,
							  method = "wilcox.test", paired = FALSE)
show_df(as.data.frame(comp_utr_len))
```

```{r utr-length-boxplot-advanced}
plot1 <- ggplot(utr_width_df, aes(x = annotation, y = log10(width))) +
	geom_violin(trim = TRUE, fill = 'lightgray', color = "lightgray") +
	geom_boxplot(width = 0.3, fill = plots_colors[1:3]) +
	stat_compare_means(comparisons = to_compare) +
	stat_compare_means(label.y = 6.5) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r utr-length-density}
# Density plot
plot1 <- ggplot(utr_width_df, aes(x = log10(width), color = annotation)) +
	geom_density() +
	scale_color_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

```{r utr-length-histogram}
# Histogram
plot1 <- ggplot(utr_width_df, aes(x = log10(width), fill = annotation)) +
	geom_histogram(bins = 30, color = "#e9ecef", alpha = 0.5, position = 'identity') +
	scale_fill_manual(values = plots_colors[1:3]) +
	labs(x = "", y = "", title = paste0("3'UTR lengths (log10) of the 3 reference annotations")) +
	theme_minimal()
plot1
```

### Venn Diagrams
Intersections counts are produced via [Intervene](https://intervene.readthedocs.io/en/latest/index.html).
This tool is based on `bedtools interset`.
We defined intersections if at least 50% of the gene from reference A overlaps a gene in reference B, and vice versa.
We also consider separately the forward and reverse strand.
The resulting euler diagram (or proportional Venn diagram) is obtained with the package `eulerr`.

```{bash, eval=FALSE, echo=FALSE}
source .rvars
sed -i '/pathOutput/c\' .rvars
echo "export pathOutput='output/01-Impact-ref-annotation-scRNA/venn_diagrams/'" >> .vars
source .rvars
mkdir -p $pathOutput
```

I retained the number of interested genes obtained with Intervene to plot the Venn diagram with `eulerr`.

```{bash, eval=FALSE, echo=FALSE}
source .rvars
echo "export genesGff1='${pathRef}galGal6.ensGene_genes.gff3'" >> .rvars
echo "export genesGff2='${pathRef}galGal6.ncbiRefSeq_genes.gff3'" >> .rvars
echo "export genesGff3='${pathRef}galGal6.refGene_fixed_genes.gff3'" >> .rvars
```

```{bash, eval=FALSE}
source .rvars

intervene venn -i $genesGff1 $genesGff2 $genesGff3 \
--figtype png \
--save-overlaps \
--bedtools-options f=0.5,r,s \
--output ${pathOutput}
```

```{r venn-refs, eval=TRUE}
venn_intersect <- euler(c("UCSC" = 248,
						  "NCBI" = 8555,
						  "Ensembl" = 9004,
						  "NCBI&Ensembl" = 8927,
						  "NCBI&UCSC" = 253,
						  "Ensembl&UCSC" = 376,
						  "NCBI&Ensembl&UCSC" = 6049))

plot(venn_intersect,
	 quantities = list(type = c("counts"),
					   col =  c("black", "black", "black"),
					   fontsize = 20),
	 labels = list(col = c("black", "black", "black"),
				   fontsize = 22),
	 edges = list(col = "white", lex = 2),
	 fills = plots_colors[3:1],
	 legend = list(side = "right"))
```

### GffCompare

```{r gffcompare-load}
gffcmp_files <- c(paste0('output/01-Impact-ref-annotation-scRNA/gffcompare/gffcmp_ensGene_', file_names[2:3],'.stats.txt'))
gffcmp_stats <- lapply(gffcmp_files, read.csv,
					   header = FALSE,
					   skip = 1,
					   sep = '\t',
					   quote = "")
gffcmp_stats[[1]]['Ref'] <- 'NCBI'
gffcmp_stats[[2]]['Ref'] <- 'UCSC'
gffcmp_df <- as.data.frame(do.call(rbind, gffcmp_stats))
show_df(gffcmp_df)
```

```{r gffcompare}
plots <- list()
plots[[1]] <- gffcmp_df %>%
	filter(V1 %in% c("Locus level")) %>%
	ggplot(aes(x = Ref, y = V2, fill = Ref)) +
	geom_bar(stat="identity") +
	scale_fill_manual(values = plots_colors[2:3]) +
	labs(x = "", y = "Percentage", title = "Sensitivity") +
	ylim(c(0, 100)) +
	guides(fill = FALSE) +
	theme_minimal()

plots[[2]] <- gffcmp_df %>%
	filter(V1 %in% c("Locus level")) %>%
	ggplot(aes(x = Ref, y = V3, fill = Ref)) +
	geom_bar(stat="identity") +
	scale_fill_manual(values = plots_colors[2:3]) +
	labs(x = "", y = "Percentage", title = "Precision") +
	ylim(c(0, 100)) +
	guides(fill = FALSE) +
	theme_minimal()

plots[[3]] <- gffcmp_df %>%
	filter(V1 %in% c("Missed loci")) %>%
	ggplot(aes(x = Ref, y = V4, fill = Ref)) +
	geom_bar(stat="identity") +
	scale_fill_manual(values = plots_colors[2:3]) +
	labs(x = "", y = "Percentage", title = "Missed loci") +
	ylim(c(0, 100)) +
	guides(fill = FALSE) +
	theme_minimal()

plots[[4]] <- gffcmp_df %>%
	filter(V1 %in% c("Novel loci")) %>%
	ggplot(aes(x = Ref, y = V4, fill = Ref)) +
	geom_bar(stat="identity") +
	scale_fill_manual(values = plots_colors[2:3]) +
	labs(x = "", y = "Percentage", title = "Novel loci") +
	ylim(c(0, 100)) +
	guides(fill = FALSE) +
	theme_minimal()

do.call(grid.arrange, plots)
```


## Impact of reference choice on scRNA-seq analyses

```{r load-sc-data}
# ReadRDS
rds_seurat <- list.files("data/rds/references/", pattern = "*obj_seurat_final.rds", full.names = TRUE)
rds_seurat <- lapply(rds_seurat, readRDS)

rds_markers <- list.files("data/rds/references/", pattern = "*markers_clustering.rds", full.names = TRUE)
rds_markers <- lapply(rds_markers, readRDS)

# Reproduce colors palette
colors_pal1 <- as.vector(yarrr::piratepal("appletv", plot.result = FALSE, trans = 0)[c(3,1,6)])
colors_pal2 <- as.vector(brewer.pal(n = 4, name = "Dark2"))
colors_pal3 <- as.vector(brewer.pal(n = 9, name = "Paired"))
colors_pal4 <- ggthemes_data[["tableau"]][["color-palettes"]][["regular"]][["Classic 20"]]
colors_pal4 <- as.vector(colors_pal4$value)
colors_pal <- list(Phase = colors_pal1,
				   Cell_type = colors_pal2,
				   Cell_subtype = colors_pal3,
				   Clusters = colors_pal4)
names(colors_pal[[1]]) <- c('G1', 'G2M', 'S')
names(colors_pal[[2]]) <- c('Progenitor', 'Neuron', 'Neural_crest', 'Mesoderm')
names(colors_pal[[3]]) <- c('RP', 'dp', 'p', 'pMN', 'p3', 'FP', 'Neuron', 'Neural_crest', 'Mesoderm')
names(colors_pal[[4]]) <- paste0('clust_', 1:20)
```

### Compare UMAP

These UMAP have been previously computed in the nextflow pipeline.
We just print them again here.

```{r dim-plots, eval=TRUE}
run_DimPlot <- function(x, val) {
	Idents(x) <- val
	DimPlot(x,
			reduction = "umap",
			dims = c(1, 2),
			label = FALSE,
			cols = colors_pal$val,
			label.size = 4,
			pt.size = 0.2)}

run_DimPlot(rds_seurat[[1]], 'Cell_type')
run_DimPlot(rds_seurat[[2]], 'Cell_type')
run_DimPlot(rds_seurat[[3]], 'Cell_type')
run_DimPlot(rds_seurat[[1]], 'Cell_subtype')
run_DimPlot(rds_seurat[[2]], 'Cell_subtype')
run_DimPlot(rds_seurat[[3]], 'Cell_subtype')

#lapply(rds_seurat, run_DimPlot, 'Cell_type')
#lapply(rds_seurat, run_DimPlot, 'Cell_subtype')
#lapply(rds_seurat, run_DimPlot, 'Phase')
#lapply(rds_seurat, run_DimPlot, 'seurat_clusters')
```

```{r feature-plots, eval=TRUE}
run_FeaturePlot <- function(x, val) {
	for (i in seq(1, length(val))) {
		if (val[i] %in% colnames(x@meta.data)) {
			print(
				FeaturePlot(x,
            		features = val[i],
            		dims = c(1, 2),
            		cols = c("grey90", brewer.pal(9,"YlGnBu")),
            		pt.size = 0.2,
            		ncol = 1) + NoLegend())}}}

run_FeaturePlot(rds_seurat[[1]], names(colors_pal[[2]]))
run_FeaturePlot(rds_seurat[[2]], names(colors_pal[[2]]))
run_FeaturePlot(rds_seurat[[3]], names(colors_pal[[2]]))

run_FeaturePlot(rds_seurat[[1]], names(colors_pal[[3]]))
run_FeaturePlot(rds_seurat[[2]], names(colors_pal[[3]]))
run_FeaturePlot(rds_seurat[[3]], names(colors_pal[[3]]))

#lapply(rds_seurat, run_FeaturePlot, names(colors_pal[[2]]))
#lapply(rds_seurat, run_FeaturePlot, names(colors_pal[[3]]))
```




### Compare proportions of populations

```{r compare-proportions}
tmp <- rds_seurat[[3]][[]]
tmp$p3 <- NA
md_cells <- rbind(rbind(rds_seurat[[1]][[]], rds_seurat[[2]][[]]), tmp)

ggplot(md_cells, aes(x = orig.ident)) +
    geom_bar(aes(fill = as.factor(Cell_type)), position = "fill") +
    theme_minimal() +
    scale_y_continuous(labels = percent) +
	scale_fill_manual(values = colors_pal$Cell_type) +
    labs(x = "", y = "", title = "Proportion of cell types", fill = "Cell type")

ggplot(md_cells, aes(x = orig.ident)) +
    geom_bar(aes(fill = as.factor(Cell_subtype)), position = "fill") +
    theme_minimal() +
    scale_y_continuous(labels = percent) +
	scale_fill_manual(values = colors_pal$Cell_subtype) +
    labs(x = "", y = "", title = "Proportion of subtypes", fill = "Cell subtype")

ggplot(md_cells, aes(x = orig.ident)) +
    geom_bar(aes(fill = as.factor(Phase)), position = "fill") +
    theme_minimal() +
    scale_y_continuous(labels = percent) +
	scale_fill_manual(values = colors_pal$Phase) +
    labs(x = "", y = "", title = "Proportion of cells in each phase of the cell cycle", fill = "Phase")

ggplot(md_cells, aes(x = orig.ident)) +
    geom_bar(aes(fill = as.factor(seurat_clusters)), position = "fill") +
    theme_minimal() +
    scale_y_continuous(labels = percent) +
	scale_fill_manual(values = as.vector(colors_pal$Clusters)) +
    labs(x = "", y = "", title = "Proportion of cells in each cluster", fill = "Cluster")
```

### Find shared differentially expressed genes

```{r de-genes-prepare}
# Prepare data to extract DE genes
create_bulk_df <- function(x) {
	tmp <- as.data.frame(cbind(x, coln = rowMeans(x)))
	tmp$gene <- rownames(tmp)
	tmp <- tmp[, c(ncol(tmp)-1,ncol(tmp))]
}
de_genes <- lapply(rds_markers, function(x) {unique(sort(x$gene))} )
raw_counts <- lapply(rds_seurat, GetAssayData, slot = "counts")
bulk_counts <- lapply(raw_counts, create_bulk_df)
colnames(bulk_counts[[1]]) <- c("Ensembl", "gene")
colnames(bulk_counts[[2]]) <- c("NCBI", "gene")
colnames(bulk_counts[[3]]) <- c("UCSC", "gene")

# Check number of DE genes with each reference
lapply(de_genes, length)

# Number of genes (all) in each reference
lapply(bulk_counts, nrow)
```

#### Between Ensembl and NCBI

We here compare DE genes.

```{r scater-DE-Ensembl-NCBI}
i <- 1
j <- 2
df_counts <- merge(bulk_counts[[i]][de_genes[[i]],],
				   bulk_counts[[j]][de_genes[[j]],])
dim(df_counts)

ref_names <- c('Ensembl', 'NCBI', 'UCSC')
df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 5 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 20, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(Ensembl), y = log10(NCBI))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)
```

We now compare all the genes.

```{r scater-all-Ensembl-NCBI}
df_counts <- merge(bulk_counts[[i]],
				   bulk_counts[[j]])
dim(df_counts)

# Plot the most divergent genes between the 2 references
df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 50 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 100, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(Ensembl), y = log10(NCBI))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)

# Plot in color the DE genes that are not shared between the 2 annotations
tmp <- de_genes[[i]]
de_excl_Ensembl <- tmp[which(!de_genes[[i]] %in% de_genes[[j]])]
length(de_excl_Ensembl)
tmp <- de_genes[[j]]
de_excl_NCBI <- tmp[which(!de_genes[[j]] %in% de_genes[[i]])]
length(de_excl_NCBI)

# DF of mutually exclusive genes
df_counts_i <- df_counts %>%
	filter(gene %in% de_excl_Ensembl) %>%
	mutate(Ref = ref_names[i])
df_counts_j <- df_counts %>%
	filter(gene %in% de_excl_NCBI) %>%
	mutate(Ref = ref_names[j])
df_counts <- rbind(df_counts_i, df_counts_j)
dim(df_counts)

colors_pal1 <- as.vector(yarrr::piratepal("info", plot.result = FALSE, trans = 0.2)[c(1,2)])
ggplot(df_counts, aes(x = log10(Ensembl), y = log10(NCBI))) +
	geom_point(aes(color = Ref)) +
	theme_minimal() +
	theme(legend.position = "none") +
	scale_color_manual(values = colors_pal1) +
	stat_cor(aes(color = Ref), method = "pearson", label.x = -2.5, size = 5)
```


#### Between Ensembl and UCSC

We here compare DE genes.

```{r scater-DE-Ensembl-UCSC}
i <- 1
j <- 3
df_counts <- merge(bulk_counts[[i]][de_genes[[i]],],
				   bulk_counts[[j]][de_genes[[j]],])
dim(df_counts)

df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 5 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 20, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(Ensembl), y = log10(UCSC))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)
```

We now compare all the genes.

```{r scater-all-Ensembl-UCSC}
df_counts <- merge(bulk_counts[[i]],
				   bulk_counts[[j]])
dim(df_counts)

# Plot the most divergent genes between the 2 references
df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 50 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 100, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(Ensembl), y = log10(UCSC))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)

# Plot in color the DE genes that are not shared between the 2 annotations
tmp <- de_genes[[i]]
de_excl_Ensembl <- tmp[which(!de_genes[[i]] %in% de_genes[[j]])]
length(de_excl_Ensembl)
tmp <- de_genes[[j]]
de_excl_UCSC <- tmp[which(!de_genes[[j]] %in% de_genes[[i]])]
length(de_excl_UCSC)

# DF of mutually exclusive genes
df_counts_i <- df_counts %>%
	filter(gene %in% de_excl_Ensembl) %>%
	mutate(Ref = ref_names[i])
df_counts_j <- df_counts %>%
	filter(gene %in% de_excl_UCSC) %>%
	mutate(Ref = ref_names[j])
df_counts <- rbind(df_counts_i, df_counts_j)
dim(df_counts)

colors_pal1 <- as.vector(yarrr::piratepal("info", plot.result = FALSE, trans = 0.2)[c(1,2)])
ggplot(df_counts, aes(x = log10(Ensembl), y = log10(UCSC))) +
	geom_point(aes(color = Ref)) +
	theme_minimal() +
	theme(legend.position = "none") +
	scale_color_manual(values = colors_pal1) +
	stat_cor(aes(color = Ref), method = "pearson", label.x = -2.5, size = 5)
```

#### Between NCBI and UCSC

We here compare DE genes.

```{r scater-DE-NCBI-UCSC}
i <- 2
j <- 3
df_counts <- merge(bulk_counts[[i]][de_genes[[i]],],
				   bulk_counts[[j]][de_genes[[j]],])
dim(df_counts)

df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 6 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 3, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(NCBI), y = log10(UCSC))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)
```

We now compare all the genes.

```{r scater-all-NCBI-UCSC}
df_counts <- merge(bulk_counts[[i]],
				   bulk_counts[[j]])
dim(df_counts)

# Plot the most divergent genes between the 2 references
df_counts$Divergent <- as.factor(ifelse(df_counts[,ref_names[i]] / df_counts[,ref_names[j]] > 50 |
	df_counts[,ref_names[j]] / df_counts[,ref_names[i]] > 100, 1, 0))
div <- df_counts[which(df_counts$Divergent == 1),]

ggplot(df_counts, aes(x = log10(NCBI), y = log10(UCSC))) +
	geom_point(aes(color = Divergent)) +
	scale_color_manual(values = c("grey", "red")) +
	theme_minimal() +
	theme(legend.position = "none") +
	geom_text_repel(data = div,
					aes(label = gene),
					size = 5,
					box.padding = unit(0.35, "lines"),
					point.padding = unit(0.3, "lines")) +
	stat_cor(method = "pearson", label.x = -2.5, size = 5)

# Plot in color the DE genes that are not shared between the 2 annotations
tmp <- de_genes[[i]]
de_excl_NCBI <- tmp[which(!de_genes[[i]] %in% de_genes[[j]])]
length(de_excl_Ensembl)
tmp <- de_genes[[j]]
de_excl_UCSC <- tmp[which(!de_genes[[j]] %in% de_genes[[i]])]
length(de_excl_UCSC)

# DF of mutually exclusive genes
df_counts_i <- df_counts %>%
	filter(gene %in% de_excl_NCBI) %>%
	mutate(Ref = ref_names[i])
df_counts_j <- df_counts %>%
	filter(gene %in% de_excl_UCSC) %>%
	mutate(Ref = ref_names[j])
df_counts <- rbind(df_counts_i, df_counts_j)
dim(df_counts)

colors_pal1 <- as.vector(yarrr::piratepal("info", plot.result = FALSE, trans = 0.2)[c(1,2)])
ggplot(df_counts, aes(x = log10(NCBI), y = log10(UCSC))) +
	geom_point(aes(color = Ref)) +
	theme_minimal() +
	theme(legend.position = "none") +
	scale_color_manual(values = colors_pal1) +
	stat_cor(aes(color = Ref), method = "pearson", label.x = -2.5, size = 5)
```



## Differences in gene length

```{r}
head(genes_width_df)
genes_width_df %>%
	mutate(diff_NCBI_Ens = NCBI - Ens) %>%
	mutate(diff_NCBI_UCSC = NCBI - UCSC) %>%
	mutate(diff_Ens_UCSC = Ens - UCSC)
head(df_counts)


```

## Differences in 3'UTR length

```{r, eval=FALSE}
head(utr_width_df)
utr_width_df %>%
	mutate(diff_NCBI_Ens = NCBI - Ens) %>%
	mutate(diff_NCBI_UCSC = NCBI - UCSC) %>%
	mutate(diff_Ens_UCSC = Ens - UCSC)

width2 <- data.frame(annotation = "NCBI", value = utr[[2]]['width'], gene = utr[[2]]['gene_name'])
colnames(width2) <- c("Ref", "length", "gene")
width3 <- data.frame(annotation = "UCSC", value = utr[[3]]['width'], gene = utr[[3]]['gene_name'])
colnames(width3) <- c("Ref", "length", "gene")
#inner_join(width2, width3, by = "gene")

# Keep unique values of 3UTR
width2 <- width2 %>% distinct() %>% group_by(gene) %>% filter(length_NCBI == max(length_NCBI))
width3 <- width3 %>% distinct() %>% group_by(gene) %>% filter(length_UCSC == max(length_UCSC))

tmp <- rbind(width2, width3)
tmp <- inner_join(width2, width3, by = "gene")
dim(utr_width_df)
head(utr_width_df)
table(utr_width_df$annotation)

tmp <- head(df_counts)

dplyr::inner_join(tmp, utr_width_df)
```



# WORK IN PROGRESS
```{r, eval=FALSE, echo=FALSE}
lapply(genes, head)
lapply(genes, nrow)

# Change geneID to geneName for Ensembl annotation
head(genes[[1]])
biomart <- read.table(file = 'biomart.txt', header = TRUE, sep = '\t')
biomart <- biomart %>% mutate(Name = Gene.stable.ID.version)
id <- match(genes[[1]]$Name, biomart$Gene.stable.ID.version)

tmp <- genes[[1]] %>%
	left_join(biomart[id, c("Name", "Gene.name")], by = "Name") %>%
    mutate(row_names = ifelse(Gene.name == "", Name, Gene.name)) %>%
    mutate(row_names_dup = ifelse(duplicated(row_names) |
								  duplicated(row_names, fromLast=TRUE ),
								  Name, row_names))

tmp[,c('Name', 'Gene.name', 'Parent', 'row_names_dup')] <- NULL
colnames(tmp)[11] <- "Name"
tmp2 <- makeGRangesFromDataFrame(tmp, keep.extra.columns=TRUE)
rtracklayer::export(tmp2, "my.gff", version = "3")

genes[[1]] <- tmp

# Prepare data for plotting
names1 <- data.frame(annotation = "Ensembl", value = genes[[1]]['Name'])
names2 <- data.frame(annotation = "NCBI", value = genes[[2]]['Name'])
names3 <- data.frame(annotation = "UCSC", value = genes[[3]]['Name'])
names_df <- rbind(names1, names2, names3)


v1 <- genes[[1]][['Name']]
v2 <- genes[[2]][['Name']]
v3 <- genes[[3]][['Name']]
tmp <- intersect(v1, v2)
tmp <- intersect(v1, v3)
tmp <- intersect(v2, v3)
# 11737 genes names in common bwt Ensembl and NCBI
# 5832 genes names in common bwt Ensembl and UCSC
# 6545 genes names in common bwt NCBI and UCSC
```

```{bash, eval= FALSE}
# with intersections

```

```{r, eval=FALSE}
file_names <- c('bedtools_ensGene_ncbiRefSeq_refGene_tab.bed',
				'bedtools_ncbiRefSeq_refGene_ensGene_tab.bed',
				'bedtools_refGene_ensGene_ncbiRefSeq_tab.bed')
ref_names <- c('Ensembl', 'NCBI', 'UCSC')
bed_files <- c(paste0('../scAnnotatiONT/tmp/convert-table/', file_names))
beds <- list()
bed <- lapply(bed_files, read.csv,
					header = FALSE,
					sep = '\t',
					quote = "")
lapply(bed, head)

tmp <- bed[[1]]
tmp$V3 <- NULL
head(tmp)
table(tmp$V1 == tmp$V4)
tmp <- unique(tmp)

tmp_head <- head(tmp, n=10)
tmp2 <- tmp_head %>% filter(V2 == "NCBI")
tmp3 <- tmp_head %>% filter(V2 == "UCSC")
full_join(tmp2, tmp3)

tmp2 %>% spread(V2, V4)

tmp <- data.frame(V1 = c('france', 'italy', 'italy', 'germany'), V2 = c('A', 'A', 'B', '.'), V4 = c('pink', 'blue', 'blue', NA))
tmp <- data.frame(V1 = c('france', 'france', 'italy', 'italy', 'germany'), V2 = c('A', 'A', 'A', 'B', '.'), V4 = c('pink', 'pink2', 'blue', 'blue', NA))
tmp %>% unique() %>% spread(V2, V4)
tmp_head %>% unique() %>% spread(V2, V4)
pivot_wider(tmp, names_from = V2, values_from = V4)
tmp <- tmp %>%
	group_by(V2) %>%
	mutate(helper = paste0("Group", row_number())) %>% 
	ungroup() %>%
	pivot_wider(names_from = V2, values_from = V4)
```





<!--chapter:end:01-Impact-ref-annotation-scRNA.Rmd-->

# Incomplete annotations induce signal loss

<!--chapter:end:02-Incomplete-annotations-induce-signal-loss.Rmd-->

# Approaches to improve transcriptome with long reads

<!--chapter:end:03-Approaches-to-improve-transcriptome-with-Long-Reads.Rmd-->

# Impact of re-annotation on scRNA-seq analyses

<!--chapter:end:04-Impact-reannotation-scRNA.Rmd-->

# Validation of novel genes with scRNA

<!--chapter:end:05-Validation-of-novel-genes-with-scRNA.Rmd-->

# A tool and pipeline to improve annotation for scRNA-seq analyses

<!--chapter:end:06-A-tool-and-pipeline-to-improve-annotation-for-scRNA.Rmd-->

# Session info
```{r session-info}
sessionInfo()
```

<!--chapter:end:07-Session-info.Rmd-->

